Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-06 paper.tex
most of beginning done
much, much cleaner

% \subsubsection{Practical Uses of Finiteness}
% Proofs of finiteness have well-known practical applications in
% constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% In section~\ref{practical}, we build a library which exploits these uses in
% Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% complex proofs over finite types.
% We frame this in terms of the principle of omniscience for finite types.
% Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% go further than the usual examples of this kind of proof automation: as well as
% proving properties about functions, we can synthesise functions whole-cloth from
% their desired properties.
% Through the unified interface for finite and countable types, we can
% reuse the automation machinery for \emph{partial} proof search over infinite
% search spaces.
% Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

We will at this point take a moment to define some of the types we used to
define \(\mathcal{E}!\).
Lists, and membership thereof, are defined using \emph{containers}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S \triangleright P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  \begin{equation}
    \begin{aligned}
      S : \text{Type}, P : S \rightarrow \text{Type} \\
      \mathbf{Container} = S \triangleright P
    \end{aligned}
  \end{equation}
  We ``interpret'' a container into a functor defined like so:
  \begin{equation}
    \llbracket S \triangleright P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}
Containers can be used to define a wide variety of functors (streams, trees,
etc.): lists are all that interest us now.
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Lists]
  The ``shape'' of lists is \(\mathbb{N}\), indicating the length of the list in
  question.
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}

@@

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  These non-recursive types have similar, simple proofs.
  For each we first provide the support list: they are \([]\), \([\text{tt}]\),
  and \([\text{false}, \text{true}]\) respectively.
  The cover proof should return an index which points at the given element: for
  \(\bot\) this function is present via the principle of explosion, for \(\top\)
  we always return a \(0\), and for \(\mathbf{Bool}\) we return a \(0\) for
  \(\text{false}\), and a \(1\) for \(\text{true}\).
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
  \end{multicols}
\end{proof}
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum) and
  Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  Both of these closures can be derived from closure of the \(\sum\) type.
  The non-dependent product is equivalent to a special case of \(\sum\):
  \begin{equation}
    A \times B \simeq \sum_{(x : A)} , B
  \end{equation}
  And disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
Where split enumerability was the enumeration form of a surjection from
\(\mathbf{Fin}\), we see here that manifest Bishop finiteness is the enumeration
form of an \emph{equivalence} with \(\mathbf{Fin}\).
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

  It suffices now to prove the following:
  \begin{equation}
    \prod_{(x : A)} , \prod_{(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).

  Formally, given a type \(A\) and a type family \(U\) on \(A\), when \(A\) is
  manifestly Bishop finite:
  \begin{equation}
    \mathcal{B}\left( A \right)
  \end{equation}
  And \(U\) is manifestly Bishop finite over all points of \(A\):
  \begin{equation}
    \prod_{(x : A)} , \mathcal{B}\left( U(x) \right)
  \end{equation}
  Then we have the following:
  \begin{equation}
    \mathcal{B}\left(\prod_{(x : A)} , U(x)\right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  The bulk of the proof now is concerned with proving the following:
  \begin{equation}
    \left( \prod_{(x : \mathbf{Fin}(n))} , \mathcal{E}!(A(x)) \right) \rightarrow \mathcal{E}!\left( \prod_{(x : \mathbf{Fin}(n))} , A(x) \right)
  \end{equation}
  Our strategy to accomplish this will be to consider functions from
  \(\mathbf{Fin}(n)\) as \(n\)-tuples over some type family \(T : \mathbb{N}
  \rightarrow \text{Type}\).
  \begin{equation}
    \begin{aligned}
      \mathbf{Tuple}(T, 0)   &= \top \\
      \mathbf{Tuple}(T, n+1) &= T(0) \times \mathbf{Tuple}(T \circ \text{suc}, n)
    \end{aligned}
  \end{equation}
  This type is manifestly Bishop finite, as it is constructed only from products
  and the unit type.

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \prod_{(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
A defining feature of split enumerability and Bishop finiteness is decidability:
both predicates imply a decidable equality function on the underlying type.
To find a predicate for finiteness which doesn't imply this, and therefore
works with types other than just sets, we \emph{truncate} the membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \sum_{(\mathit{xs} : \mathbf{List}(A))} , \prod_{(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\subsection{Propositional Truncation}
\begin{rm-definition}[Propositional Truncation]
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\).
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

  It is defined as a Higher Inductive Type:
  \begin{equation}
    \begin{aligned}
      \lVert A \rVert &= \\
        &| \; \lvert \cdot \rvert : A \rightarrow \lVert A \rVert ; \\
        &| \; \text{squash} : \prod_{(x, y : \lVert A \rVert)} , x \equiv y  ; \\
    \end{aligned}
  \end{equation}

  To eliminate from \(\lVert A \rVert\) we need to prove that we don't ``care''
  about the specific choice of \(A\).
  In other words, given a function \(f : A \rightarrow B\), to construct a
  function \(\lVert A \rVert \rightarrow B\), \(f\) needs to satisfy the following:
  \begin{equation}
    \prod_{(x , y : A)} , f(x) \equiv f(y)
  \end{equation}
  Formally speaking, \(f\) needs to be ``coherently constant''
  \cite{krausGeneralUniversalProperty2015}.
\end{rm-definition}
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).

  The support list will be a singleton list containing the single point in the
  type, and the cover proof will be an index pointing at the first element in
  the support list.

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestEnumerable.tex]{circ-inst}
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
Deriving a surjection from \(\mathbf{Fin}\) is straightforward:
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}
\subsection{Relation to Split Enumerability}
As split enumerability is stronger than manifest enumerability, we can convert
from one to the other easily.
\begin{rm-lemma}
  Any split enumerable type is also manifestly enumerable.
\end{rm-lemma}
\begin{proof}
  The proof carries over via truncation of the cover proof.
\end{proof}
The main missing piece from manifest enumerability is decidable equality: by
truncating the membership proof, we have removed the ability to distinguish
(decidably) between members of the type.
Indeed, resupplying decidable equality is precisely sufficient to recover split
enumerability.

\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

  For the cover proof, we need a way to get the contents out of a truncated
  value.
  We can do exactly that with the following lemma, called recompute:
  Given a decision procedure for some type \(A\), and a truncation for the
  proposition \(A\), we can discount the possibility of \(A\) being false, and
  therefore extract the true decision.

  For the cover proof, our obligation now becomes constructing a decision for
  membership of the support list.
  This is straightforward given decidable equality.

>>>>

% \subsubsection{Practical Uses of Finiteness}
% Proofs of finiteness have well-known practical applications in
% constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% In section~\ref{practical}, we build a library which exploits these uses in
% Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% complex proofs over finite types.
% We frame this in terms of the principle of omniscience for finite types.
% Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% go further than the usual examples of this kind of proof automation: as well as
% proving properties about functions, we can synthesise functions whole-cloth from
% their desired properties.
% Through the unified interface for finite and countable types, we can
% reuse the automation machinery for \emph{partial} proof search over infinite
% search spaces.
% Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

We will at this point take a moment to define some of the types we used to
define \(\mathcal{E}!\).
Lists, and membership thereof, are defined using \emph{containers}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S \triangleright P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  \begin{equation}
    \begin{aligned}
      S : \text{Type}, P : S \rightarrow \text{Type} \\
      \mathbf{Container} = S \triangleright P
    \end{aligned}
  \end{equation}
  We ``interpret'' a container into a functor defined like so:
  \begin{equation}
    \llbracket S \triangleright P \rrbracket = \Pi {(X : \text{Type})} , \Sigma {(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}
Containers can be used to define a wide variety of functors (streams, trees,
etc.): lists are all that interest us now.
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Lists]
  The ``shape'' of lists is \(\mathbb{N}\), indicating the length of the list in
  question.
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}

@@

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  These non-recursive types have similar, simple proofs.
  For each we first provide the support list: they are \([]\), \([\text{tt}]\),
  and \([\text{false}, \text{true}]\) respectively.
  The cover proof should return an index which points at the given element: for
  \(\bot\) this function is present via the principle of explosion, for \(\top\)
  we always return a \(0\), and for \(\mathbf{Bool}\) we return a \(0\) for
  \(\text{false}\), and a \(1\) for \(\text{true}\).
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
  \end{multicols}
\end{proof}
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \Pi {(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum) and
  Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  Both of these closures can be derived from closure of the \(\sum\) type.
  The non-dependent product is equivalent to a special case of \(\sum\):
  \begin{equation}
    A \times B \simeq \Sigma {(x : A)} , B
  \end{equation}
  And disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
Where split enumerability was the enumeration form of a surjection from
\(\mathbf{Fin}\), we see here that manifest Bishop finiteness is the enumeration
form of an \emph{equivalence} with \(\mathbf{Fin}\).
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

  It suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).

  Formally, given a type \(A\) and a type family \(U\) on \(A\), when \(A\) is
  manifestly Bishop finite:
  \begin{equation}
    \mathcal{B}\left( A \right)
  \end{equation}
  And \(U\) is manifestly Bishop finite over all points of \(A\):
  \begin{equation}
    \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
  \end{equation}
  Then we have the following:
  \begin{equation}
    \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  The bulk of the proof now is concerned with proving the following:
  \begin{equation}
    \left( \Pi {(x : \mathbf{Fin}(n))} , \mathcal{E}!(A(x)) \right) \rightarrow \mathcal{E}!\left( \Pi {(x : \mathbf{Fin}(n))} , A(x) \right)
  \end{equation}
  Our strategy to accomplish this will be to consider functions from
  \(\mathbf{Fin}(n)\) as \(n\)-tuples over some type family \(T : \mathbb{N}
  \rightarrow \text{Type}\).
  \begin{equation}
    \begin{aligned}
      \mathbf{Tuple}(T, 0)   &= \top \\
      \mathbf{Tuple}(T, n+1) &= T(0) \times \mathbf{Tuple}(T \circ \text{suc}, n)
    \end{aligned}
  \end{equation}
  This type is manifestly Bishop finite, as it is constructed only from products
  and the unit type.

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
A defining feature of split enumerability and Bishop finiteness is decidability:
both predicates imply a decidable equality function on the underlying type.
To find a predicate for finiteness which doesn't imply this, and therefore
works with types other than just sets, we \emph{truncate} the membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Propositional Truncation]a \label{prop-trunc}
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\).
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

  It is defined as a Higher Inductive Type:
  \begin{equation}
    \begin{aligned}
      \lVert A \rVert &= \\
        &| \; \lvert \cdot \rvert : A \rightarrow \lVert A \rVert ; \\
        &| \; \text{squash} : \Pi {(x, y : \lVert A \rVert)} , x \equiv y  ; \\
    \end{aligned}
  \end{equation}

  We will use and consume values of the type \(\lVert A \rVert\) in three main
  ways.
  \begin{enumerate}
    \item \label{elim-prop-prop} We can first eliminate from \(\lVert A \rVert\)
      into any type which is a proposition.
      In other words, given a function \(f : A \rightarrow B\), and a proof that
      \(B\) is a proposition, we can construct a function \(\lVert A \rVert
      \rightarrow B\).
    \item \label{elim-prop-monad} As a consequence of this first point, we can
      always eliminate into another propositionally truncated type.
      As a result, \(\lVert \cdot \rVert\) forms a Monad: for our purposes, this
      simply means tat we can work ``under'' a propositional truncation in an
      ergonomic way.
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\).
      \begin{equation}
        \Pi {(x , y : A)} , f(x) \equiv f(y)
      \end{equation}
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestEnumerable.tex]{circ-inst}
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  As with the other surjection proof (lemma~\ref{split-enum-surj}), this is
  simply a reassociation.
\end{proof}
\subsection{Relation to Split Enumerability}
\begin{rm-lemma}
  Any split enumerable type is also manifestly enumerable.
\end{rm-lemma}
\begin{proof}
  The proof carries over via truncation of the cover proof.
\end{proof}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

  For the cover proof, we need a way to get the contents out of a truncated
  value.
  We can do exactly that with the following lemma, called recompute:
  given a decision procedure for some type \(A\), and a truncation for the
  proposition \(A\), we can discount the possibility of \(A\) being false, and
  therefore extract the true decision.
